\ourtitle

### 2020-05-18 paper/article.tex
Add sections about grids and keyboards to the paper

\def\ourtitle{Semantics of Implicitly Timed Musical Events}
\title{\ourtitle}
\titlerunning{\ourtitle}

@@

\funding{This research was partially funded by Portuguese National funds
(PIDDAC), through the FCT – Fundação para a Ciência e Tecnologia and
FCT/MCTES under the scope of the projects UIDB/05549/2020 and
UIDB/00319/2020.  Bernardo Sacanene acknowledges from the Angolan
govenment his PhD grant, through INAGBE (Instituto Nacional de Gestão
de Bolsas de Estudos).}

@@

\section{Introduction}
Programming languages are sometimes described as data plus code. Some of that data is ingested through IO (like keyboards, files, sockets and others). Some of that data is generated by algorithms at runtime. But there's also a third category that blurs the line between data and code: literal values.

Virtually every popular language nowadays has custom syntax to allow the programmer to describe some very common (and primitive) data types, such as numbers, booleans or strings. Many modern languages even have syntax for more advanced data structures, such as arrays and dictionaries (or hash tables).

When we cross into the territory of DSLs \textit{(Domain Specific Languages)}, there is an even bigger multitude of custom syntax for the most varied data types. In this paper we will discuss one possible way of of describing musical events (such as notes, chords or rests), and more importantly, musical arrangements and melodies. We will see how to treat these structures as data values that can be integrated into a programming language.

Such problem can be divided into two parts: the syntax used for describing the notes and the operators the compose them; and the semantics of the generated events, how they are stored in memory, and how their temporal properties are handled without forcing the programmer/user to manually type them. The former is a relatively easy problem, and to minimize the learning curve for new users, we adopted a simplified version of the very popular note declaration syntax from the ABC Notation project\cite{AbcNotation}, with some minor changes. The latter is the one we'll be discussing in greater detail.

The simplest way of generating such musical events is to use already common, \textit{low-level}, programming mechanisms, such as using a procedural approach where the user creates each event manually by calling a function and providing as parameters all the events' information, such as it's timestamp and duration. This is the approach used by some of the existing languages in this space, such as \textit{SonicPi}.

\begin{lstlisting}[caption={Example of a hypothetical imperative API for creating events},label=list:8-6,captionpos=t,abovecaptionskip=-\medskipamount]
play_note( 0, 100, 'A' );
play_note( 100, 50, 'B' );
play_note( 150, 200, 'C' );
\end{lstlisting}

Instead we will take a look at a more \textit{functional} approach with custom syntax and operators. Musical events are treated as sequences, anc as such can be stored in variables, passed around and trasformed. For musical events, we will be exploring a way to define them in code, as \textit{musical literals}.

\begin{lstlisting}[caption={Our proposed declarative syntax that calculates timings implicitly},label=list:8-6,captionpos=t,abovecaptionskip=-\medskipamount]
play( A B/2 C2 );
\end{lstlisting}

But more that just being able to define those events, we will also be interested in exploring how well they integrate with existing and common programming language constructs, like variables, functions, loops and other control structures.

@@

The first requirement, \textbf{interactivity}, relates to our goal of not only being able to generate music offline, but also in a live environment: give the user the ability to program several snippets of musical events, and then control them through a virtual keyboard or through other interactive means.

The second requirement, \textbf{laziness}, refers to a concept that is familiar in functional programming languages: values are generated when we need them, not earlier. In our case, this implies that a musical sequence could be potentially infinite (like an infinite repetition of some arrangement). If playing this music live, the musician could determine to stop this arrangement sooner or later.

Given these two requirements, we can conclude we \textbf{cannot} generate all music events at the start and then sort them to play them in order. We conclude that the events must always be sorted already.

\begin{lemma}[Total Order]
\label{lemma:total-order} All operators must return a sequence of events in respecting our time unit's total order.
\end{lemma}

\paragraph*{Data Model}
The basic premise is that expressions can generate a special data type: \textbf{Music}. Music is simply a sequence of ordered musical events.

A musical \textbf{Event} can be one of many things, such as a \textit{note}, a \textit{chord}, or even more implementation-specific events like MIDI messages. While all events must have a start time, some events can be instantaneous (events with a duration of zero).

@@

\begin{description}
    \item[Concatenation] \verb|Music1 Music2 ... MusicN|
    \item[Parallel] \verb'Music1 | Music2 | ... | MusicN'
    \item[Repetition] \verb'Music * Integer'
    \item[Arpegio] \verb'Chord * Music'
    \item[Transpose] \verb'Music + Integer' \textit{and} \verb'Music - Integer'
\end{description}

@@

@@

\begin{lstlisting}[caption={Creating a Note Event},label=list:8-6,captionpos=t,abovecaptionskip=-\medskipamount]
c'1/4
\end{lstlisting}

\paragraph*{Concatenation}
We've seen how single events' creation is handled. Now it is important for us to see how we can combine those events together. And probably the most straightforward operator of all, concatenation, it simply consumes each event. Each event, as we've seen before, is responsible for seeking the context depending on the event's duration.
\begin{lstlisting}[caption={Snippet of Wet Hands by C418},label=list:8-6,captionpos=t,abovecaptionskip=-\medskipamount]
S4/4 T74 L/8 V90;
A, E A B ^c B A E D ^F ^c e ^c A3;
\end{lstlisting}

\begin{figure}[h]
  \centering
  {%
  \setlength{\fboxsep}{0pt}%
  \setlength{\fboxrule}{0pt}%
  \fbox{\includegraphics[width=.85\linewidth]{../code/musikla/musikla/examples/paper/concatenation2.png}}%
  }%
  \caption{Generated music sheet for concatenation\protect\footnotemark.}
  \label{fig:concatenation}
\end{figure}

@@

\begin{lstlisting}[caption={Intro to Westworld's Theme by Ramin Djawadi},label=list:8-6,captionpos=t,abovecaptionskip=-\medskipamount]
I1 S6/8 T140 L/8 V90;
A*11 G F*12
\end{lstlisting}

\begin{figure}[h]
  \centering
  {%
  \setlength{\fboxsep}{0pt}%
  \setlength{\fboxrule}{0pt}%
  \fbox{\includegraphics[width=.85\linewidth]{../code/musikla/musikla/examples/paper/repetition2.png}}%
  }%
  \caption{Generated music sheet for repetition.}
  \label{fig:repetition}
\end{figure}

@@

\begin{lstlisting}[caption={Snippet of Soft to Be Strong by Marina},label=list:8-6,captionpos=t,abovecaptionskip=-\medskipamount]
T120 V70 L1;
[^Cm] [BM] | r/4 ^g/4 ^g/4 ^f/2 e/8 ^d3/8 ^c
\end{lstlisting}

\begin{figure}[h]
  \centering
  {%
  \setlength{\fboxsep}{0pt}%
  \setlength{\fboxrule}{0pt}%
  \fbox{\includegraphics[width=.85\linewidth]{../code/musikla/musikla/examples/paper/parallel.png}}%
  }%
  \caption{Generated music sheet for parallel}
  \label{fig:parallel}
\end{figure}

@@

\paragraph*{Variables and Functions}
Up until now we've analyzed situations where the timing of the events is known at the moment of their creation (through inspection of the context passed onto them). However, there can be situations where the events have to be created before their time location is known.

We decided to take an approach to integrate this type of functionality that does not require different semantics for creating events \textit{"live"} and for storing events in all operators. Instead, the only changes occur in variable declarations and when inserting variables in the middle of other expressions.

When declaring a variable, its expression is evaluated with custom context, forked from the main context, with the cursor reset back to zero. An important thing to note is that even though we are storing the music in a variable, we still take care to respect the principle of laziness present throughout our language: when the variable is declared, no actual events are created. Later, whenever the variable is used, and an event is requested of the variable, it relays that request to the expression responsible for providing the event.

However, one important caveat is that the variable cannot simply emit the events as they are created, because as we mentioned before, their context is a different one from the main context (with its cursor set to zero at the start of the expression).

Let's assume that a variable $C$ holds the value of the \texttt{cursor(ctx)}, it \texttt{ctx} is the context where the variable is being evaluated (not declared). Then, each event would be cloned and its starting time would have the constant $C$ added to it.

This raises an important detail that we must not forget: many of the musical events are created and then emitted directly. However, as we've seen, some can be stored in variables. And variables can be used in many ways, and the same variable can be used multiple times. This means that if one were to mutate an event (by change its start time or its duration, for instance), we would be changing the the contents of the variable. This could be an acceptable approach. But we have opted instead for a more functional approach, treating those musical events as more akin to primitive values (as we would treat numbers or event strings in modern languages). This means that when we change the start time of an event, we are actually cloning it.

This works well enough because those events are very lightweight objects, and the benefits of not having their values mysteriously changed outweight the small cost of a possible unnecessary allocation, in case that event was actually only used once.

\paragraph*{Functions}
When designing functions into our language, we decided to keep the semantics simple. Emitting events inside a function is similar to its return value being an iterator that gives out the emitted events on demand. This means that a value cannot both emit musical events, while also returning other values manually through a return statement.

@@

In terms of managing the implicit context, functions are treated in very similar ways to the regular operators. They receive the context that was active at their call site and are evaluated with it. This means that any events emitted will have begin at whatever time the context's cursor marks.

Their arguments, however, are treated the same way as variable declarations (which is what they really are), meaning that each argument gets a custom fork of the call site's context with its \texttt{cursor} set to zero.

Here we can see a small snippet of the beginning of Fugue 2 in C minor in Book I of the J.S. Bach’s Well-Tempered Clavier, and how using functions can help us visualize the structure behind music.

\begin{lstlisting}[caption={Example of repeating the same note},label=list:8-6,captionpos=t,abovecaptionskip=-\medskipamount]
fun fugue ( $subj, $resp ) => 
    ( $subj $resp | stretch( r, $subj ) ( $subj + 7 ) );

@@

\begin{figure}[h]
  \centering
  {%
  \setlength{\fboxsep}{0pt}%
  \setlength{\fboxrule}{0pt}%
  \fbox{\includegraphics[width=.85\linewidth]{../code/musikla/musikla/examples/paper/fugue.png}}%
  }%
  \caption{Generated music sheet for fugue example}
  \label{fig:parallel}
\end{figure}
% https://drive.google.com/open?id=1-igVwdhXMxOSE1cMg_8HimDOXMhsY9US

>>>>

\def\ourtitle{Musikla: Language for Generating Musical Events}
\title{\ourtitle}
\titlerunning{\ourtitle}

@@

%\funding{This research was partially funded by Portuguese National funds
%(PIDDAC), through the FCT – Fundação para a Ciência e Tecnologia and
%FCT/MCTES under the scope of the projects UIDB/05549/2020 and
%UIDB/00319/2020.  Bernardo Sacanene acknowledges from the Angolan
%govenment his PhD grant, through INAGBE (Instituto Nacional de Gestão
%de Bolsas de Estudos).}

@@

\section{Introduction}
Musikla stands for Music and Keyboard Language. Our goal is to develop a DSL (\textit{Domain Specific Language}) that allows treating musical events as regular data in a programming language. More than generating these musical events offline, we want to be able to easily declare keyboards that map keys and expressions to mutate the state or play musical events (or both).

The project can be sectioned in three different, modular layers: inputs, the language, and outputs. While music events can be described as code literals in our language, they can also originate from many other sources. After being processed by our language, it then emits a stream of musical events to the \textbf{Player}, which then multiplexes those events into however many outputs the user defined.

% TODO Translate diagram to english

\begin{figure}[ht]
  \centering
  {%
  \setlength{\fboxsep}{0pt}%
  \setlength{\fboxrule}{0pt}%
  \fbox{\includegraphics[width=.65\linewidth]{../dissertation/rpd/img/diagram_architecture_en_vertical.png}}%
  }%
  \caption{The three main layers of the project.}
  \label{fig:architecture}
\end{figure}

While the development of both the input and output layers, as well as their many respective components presents by itself many interesting challenges that could be discussed, we will instead focus this paper on the aspects of the middle layer: the \textit{interpreter}, while ackowledging the existence (and their effects) of the layers that wrap around it.

As such, the problem of developing the interpreter can be divided into two parts: the syntax used for describing the notes and the operators the compose them inside the language; and the semantics of the generated events, how they are stored in memory, and how their temporal properties are handled without forcing the programmer/user to manually type them.

Designing the syntax for describing those musical expressions, especially given our strong desire to make those musical expressions first class citizens, like other primitive data types in most programming languages (such as numbers, strings or arrays are), did unearth some challenges. To minimize the learning curve for new users, and avoid reinventing the wheel, we decided to adopt a subset of the very popular note declaration syntax from the ABC Notation project\cite{AbcNotation} and integrate it with our language.

As for the execution model, we decided to go with a tree-walker interpreter\cite{CraftingInterpreters}. Altough computationally slower than other alternatives (such as a bytecode virtual machine), the ease of implementation allowed us to prototype and develop features extremely fast. And with a more mature and stable language in the future, there is always the potential to rewrite the interpreter if performance or latency ever reveal themselves as potential problems.

The simplest way of generating such musical events in a programming language is to use already common, \textit{low-level}, programming mechanisms, such as using a procedural approach where the user creates each event manually by calling a function and providing as parameters all the events' information, such as it's timestamp and duration. This is the approach used by some of the existing languages in this space, such as \textit{SonicPi}.

\begin{lstlisting}[caption={Example of a hypothetical imperative API for creating events},label=list:1,captionpos=t,abovecaptionskip=-\medskipamount]
play_note( 0, 100, 'A' );
play_note( 100, 50, 'B' );
play_note( 150, 200, 'C' );
\end{lstlisting}

Instead we decided to follow a more \textit{functional} approach, with custom syntax and operators, as well as the hability to describe those events in a single expression. Musical events are treated as sequences, and as such can be stored in variables, passed around inside functions and trasformed. So, for musical events, we will be exploring a way to define them in code, as \textit{musical literals}.

\begin{lstlisting}[caption={Our proposed declarative syntax that calculates timings implicitly},label=list:2,captionpos=t,abovecaptionskip=-\medskipamount]
play( A B/2 C2 );
\end{lstlisting}

@@

The first requirement, \textbf{Interactivity}, relates to our goal of not only being able to generate music offline, but also in a live environment: give the user the ability to program several snippets of musical events, and then control them through a virtual keyboard or through other interactive means.

The second requirement, \textbf{Laziness}, refers to a concept that is familiar in functional programming languages: values are generated when we need them, not earlier. In our case, this implies that a musical sequence could be potentially infinite (like an infinite repetition of some arrangement). If playing this music live, the musician could determine to stop this arrangement sooner or later.

Given these two requirements, we can conclude we \textbf{cannot} generate all music events at the start and then sort them to play them in order. Because of that, the events must always be sorted already.

\begin{lemma}[Total Order]
\label{lemma:total-order} All operators must return a sequence of events that respects our time unit's total order.
\end{lemma}

\paragraph*{Data Model}
The basic premise is that expressions can generate a special data type: \textbf{Music}. Music is simply a sequence (or stream) of ordered musical events.

A musical \textbf{Event} can be one of many things, such as a \textit{note}, a \textit{chord}, or even more implementation-specific events like MIDI messages. While all events must have a start time, some events can be instantaneous (events with a duration of zero time units).

@@

\begin{description}
    \item[Concatenation] \verb|Music1 Music2 ... MusicN|

        \textbf{type} \verb|List[Music] -> Music|
    \item[Parallel] \verb'Music1 | Music2 | ... | MusicN'

        \textbf{type} \verb|List[Music] -> Music|
    \item[Repetition] \verb'Music * Integer'

        \textbf{type} \verb|Music, Integer -> Music|
    \item[Arpegio] \verb'Chord * Music'

        \textbf{type} \verb|Chord, Music -> Music|
    \item[Transpose] \verb'Music + Integer' \textit{and} \verb'Music - Integer'

        \textbf{type} \verb|Music, Integer -> Music|
\end{description}

@@

\paragraph*{Grids}
Another type available in our language are grids. Also known in most music applications as the process of quantization \cite{Quantization}. The reason it is so useful in our language is that when receiving input as musical events from a live keyboard, their timings are naturally more prone to having small discrepancies that can become more apparent when we then mix them with generated musical events (which have precise timings).

Having events always aligned with such a grid can also make computations and transformations of such events easier and simpler, which is always a plus for our language.

\begin{description}
    \item[Create a grid] \verb|Grid(size)|

        \textbf{type} \verb|Fraction -> Grid|
    \item[Aligning] \verb'grid::align(music)'

        \textbf{type} \verb|Grid, Music -> Music|
    \item[Compose Grids] \verb'Grid::compose(grid1, grid2, ..., gridN)'

        \textbf{type} \verb|List[Grid] -> Grid|
\end{description}

We can see that apart from the basic operations of creating a grid and aligning events to said grid, we also want the ability to compose multiple grids (of different precisions). We will approach this matter in more detail later.

\paragraph*{Keyboards}
A core part of the language is our hability to declare keyboards, which we can describe as mappings between \texttt{Keys} and \textit{Musical Expressions}.

Each expression can mutate the state (changing variables or calling functions), return some music (sequence of musical events) to be played, or both.

Some of the operations we want to be able to perform with keyboards are as follows:

\begin{description}
    \item[Create a keyboard] \verb|keyboards\create()|

        \textbf{type} \verb|() -> Keyboard|
    \item[Binding a Key] \verb'keyboard::register(key, expression)'

        \textbf{type} \verb|Keyboard, Key, Expression -> Keyboard|
    \item[Mapping a keyboard] \verb'keyboard::map(transformer)'

        \textbf{type} \verb|Keyboard, ( Music -> Music ) -> Keyboard|
    \item[Aligning with a Grid] \verb'keyboard::with_grid(grid)'

        \textbf{type} \verb|Keyboard, Grid -> Keyboard|    
\end{description}

@@

\begin{lstlisting}[caption={Creating a Note Event},label=list:3,captionpos=t,abovecaptionskip=-\medskipamount]
c'1/4
\end{lstlisting}

\paragraph*{Concatenation}
We've seen how single events' creation is handled. Now it is important for us to see how we can combine those events together. And probably the most straightforward operator of all, concatenation, it simply consumes each event. Each event, as we've seen before, is responsible for seeking the context depending on the event's duration.
\begin{lstlisting}[caption={Snippet of Wet Hands by C418},label=list:4,captionpos=t,abovecaptionskip=-\medskipamount]
S4/4 T74 L/8 V90;
A, E A B ^c B A E D ^F ^c e ^c A3;
\end{lstlisting}

\begin{figure}[ht]
  \centering
  {%
  \setlength{\fboxsep}{0pt}%
  \setlength{\fboxrule}{0pt}%
  \fbox{\includegraphics[width=.85\linewidth]{../code/musikla/musikla/examples/paper/concatenation2.png}}%
  }%
  \caption{Generated music sheet for concatenation\protect\footnotemark.}
  \label{fig:concatenation}
\end{figure}

@@

\begin{lstlisting}[caption={Intro to Westworld's Theme by Ramin Djawadi},label=list:5,captionpos=t,abovecaptionskip=-\medskipamount]
I1 S6/8 T140 L/8 V90;
A*11 G F*12
\end{lstlisting}

\begin{figure}[ht]
  \centering
  {%
  \setlength{\fboxsep}{0pt}%
  \setlength{\fboxrule}{0pt}%
  \fbox{\includegraphics[width=.85\linewidth]{../code/musikla/musikla/examples/paper/repetition2.png}}%
  }%
  \caption{Generated music sheet for repetition.}
  \label{fig:repetition}
\end{figure}

@@

\begin{lstlisting}[caption={Snippet of Soft to Be Strong by Marina},label=list:6,captionpos=t,abovecaptionskip=-\medskipamount]
T120 V70 L1;
r/4 ^g/4 ^g/4 ^g/4   ^f/2 e/8 ^d3/8   ^c2 | [^Cm] [BM]  [AM] [BM] 
\end{lstlisting}

\begin{figure}[ht]
  \centering
  {%
  \setlength{\fboxsep}{0pt}%
  \setlength{\fboxrule}{0pt}%
  \fbox{\includegraphics[width=.85\linewidth]{../code/musikla/musikla/examples/paper/parallel2.png}}%
  }%
  \caption{Generated music sheet for parallel}
  \label{fig:parallel}
\end{figure}

@@

\paragraph*{Variables and Functions}
All expressions that are assigned to a variable run in a forked context, with it's cursor set to zero initially. Musical expressions inside variables are still lazy (meaning they only calculate each musical event when the variable is first used, not declared) but the events are cached to prevent the calculations from being performed every time. This cache is then garbage collected when the variable is no longer in use.

Since events declared inside variables have their start time set relative to zero, it needs to be calculated each time the variable is used to replace it with the correct value. This highlights an important aspect of our language: the need for each musical event to be immutable, and any changes made to them to actually be implemented as new instances of the event.

This works well enough because those events are very lightweight objects, and the benefits of not having their values mysteriously changed midway during execution outweight the small cost of a possible unnecessary allocation of an event that would only be used in one place instead of many.

Function calls, on the other hand, pass the current context to the inside of the function, so  that any events played there now their correct times.

When integrating functions into our language, we decided to keep the semantics simple. Emitting musical events inside a function is similar to its return value being an iterator that gives out the emitted events on demand. This means that a value cannot both emit musical events, while also returning other values manually through a return statement.

@@

Here we can see a small snippet of the beginning of \textit{Fugue 2 in C minor in Book I of the J.S. Bach’s Well-Tempered Clavier}, and how using functions and variables can help us visualize the structure behind music.

\begin{lstlisting}[caption={Example of repeating the same note},label=list:7,captionpos=t,abovecaptionskip=-\medskipamount]
fun fugue ( $subj, $resp ) => 
    ( $subj $resp | stretch( r, $subj ) ( $subj + 7 ) );

@@

\begin{figure}[ht]
  \centering
  {%
  \setlength{\fboxsep}{0pt}%
  \setlength{\fboxrule}{0pt}%
  \fbox{\includegraphics[width=.85\linewidth]{../code/musikla/musikla/examples/paper/fugue.png}}%
  }%
  \caption{Generated music sheet for fugue example}
  \label{fig:fugue}
\end{figure}
% https://drive.google.com/open?id=1-igVwdhXMxOSE1cMg_8HimDOXMhsY9US

\paragraph*{Grids}
To define a grid there is only one parameter required: the length of it's cells. When aligning musical events, anything inside each cell will be pushed to the closest edge of the cell.

Grids are highly customizable too, however. They have multiple parameters, such as \texttt{forgiveness} and \texttt{range}, that determine when an event is affected by the grid (depending on how close it's start time is to the edge of the cell). Each parameter can even be customized separately for the left and right sides of the cell's edge.

Let's take a look at an example of a grid. In this example the grid has a cell size of \textbf{1}. We define the same values for both left and right sides just for the sake of this demonstration, but each side could have different values.

\begin{lstlisting}[caption={Declaring a grid},label=list:8,captionpos=t,abovecaptionskip=-\medskipamount]
$grid = Grid( 1,
    forgiveness_left = 125, # 1/8
    range_left = 375, # 3/8
    forgiveness_right = 125, #1/8
    range_right = 375 # 3/8
);
\end{lstlisting}

\begin{figure}[ht]
  \centering
  {%
  \setlength{\fboxsep}{0pt}%
  \setlength{\fboxrule}{0pt}%
  \fbox{\includegraphics[width=.95\linewidth]{./img/grids.png}}%
  }%
  \caption{Representation of two cells from this grid.}
  \label{fig:grid}
\end{figure}

We can see in this timeline two cells (each with a size of 1). Any events that fall in the yellow and grey areas are ignored (meaning their timestamps are not changed) while events in the green areas are pushed to whatever edge is closest. But even this behavior can be customized, forcing events to always go to the previous edge cell, or always to the next.

It is then trivial to see how we could compose multiple grids in sequence, each with different ranges (green areas) that capture different events and align them accordingly.

\paragraph*{Keyboards}
Finally we can combine all the systems we've described above, from musical expressions, grids, variables and functions, and devise a compact way of describing virtual keyboards.

To make the process of designing keyboards less verbose, we've added syntactic sugar to this process, that is translated in the background to regular function calls registering each key binding.

While a picture maybe worth a thousand words, a good example is worth maybe even more. So here we can take a brief look at the workflow for defining two keyboards (that are active at the same time). The first keyboard has all the musical keys (the chords and single notes we want), all aligned by a custom grid.

The second keyboard binds to the up and down arrow keys and allow us to change the virtual instrument through which we play the sounds of the notes in the keyboard (those instruments can be identified by an integer and usually follow the General MIDI standard\cite{GeneralMIDI}).

\begin{lstlisting}[caption={Creating a keyboard that can play multiple instruments},label=list:9,captionpos=t,abovecaptionskip=-\medskipamount]
$inst = 0;

fun spin_instrument ( ref $instrument, $change ) {
    $instrument = $instrument + $change;

    setinstrument( $instrument );
};

@keyboard {
    a: [^Cm];   s: [BM];    d: [AM];    f: [EM];    g: [^Fm];
    1: ^c;      2: ^d;      3: e;       4: ^f;      5: ^g;
    6: b;       7: ^c';     8: ^d';     9: e';
}::with_grid( Grid( 1 / 16 ) );

@keyboard {
    up: spin_instrument( $inst, 1 );
    down: spin_instrument( $inst, -1 );
};
\end{lstlisting}

Keyboards are objects (that we could save in a variable for example) and that can perform many operations, like unions and intersections, or maps and filters. They can be enabled and disabled at runtime, and their keys can be simulated to be pressed and released.

More than that, we don't need to restrict ourselves to computer keyboards. We can for instance, define bindings between MIDI events and musical expressions, so that when we connect a piano keyboard to our computer, we can use each piano key to play more than a single note.

Since like we've seen keyboard keys are not limited to computer keyboards, we can imagine the possibilities of events we could listen to: knobs, mouse buttons, the mouse scroll wheel. We could even create an event that could, for example, listen on a socket and trigger when a message is received, allowing in that way our musical applications to be controlled remotely.

The result is that our keyboards are extremely extensible and allow for a great deal of creativity. And thanks to our tight integration with the Python language, those extensions can be easily integrated and don't require hacking the source code or recompiling the application.
